Assessment of forestry road network projection methods using the “roads” package in R

### 2022-07-18 analysis/paper/paper.Rmd
formatting

(Introduction round 2: road context)
Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use. These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network. These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes. On the other hand, many spatial stochastic forest management simulations haven’t included forest road development at all (citations). One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes. As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).  
The goal of our model is to provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations, and can also be applied as an element of more complex models. Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes. Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network. In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially. 
In the first section of the paper (methods), we introduce the study area used to assess the accuracy of the projection, we describe the forest road projection methods, the data required, and the various metrics used in order to analyze the similarities and differences between the projected and observed forest road networks across the landscape. Next, we present and discuss the results of our projections across the landscape and compare the metrics spatially and aspatially.

@@

## Road network projection methods
Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We used three methods for solving the MTAP in the roads package. These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST). In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.
The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959). Dijkstra’s algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section “weighted graph”). Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction. This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.
The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph. The minimum spanning tree was solved using Kruskal’s algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop. This method was most similar to Clark et al. (2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling. An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths. 
        A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods. All the default settings such as skidding distance, etc. were left as they were within the QGIS plugin.

@@

Table 1. Definitions of road building costs for building the cost surface. 
Cost Type
	Road Attribute
	Value ($ per km)
	Source
	Barriers of road development
	Major lakes
	infinity
	Major water bodies of BC
	Impedance
	Existing roads
	0.01
	Assumed
	Road Construction Cost Structure

	Pipe crossing
	1911
	Interior Appraisal Manual
	River crossing
	2130
	Interior Appraisal Manual
	Base cost
	16178 + 504*slope
	Interior Appraisal Manual

The road network used was a combination of the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) datasets from the Government of British Columbia’s Data Catalogue. The existing road network inputted into the model was created by filtering out the forest tenure roads with an award date greater than 1990. Because the objective of the roads package is to project forest resource roads, the DRA roads were left untouched for the existing 1990 road network. The projection results would be compared with the observed modern road network which was simply a complete and unfiltered combination of the DRA and Forest Tenure road datasets, representing the most complete and up-to-date road network for the area.   
Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) Government of British Columbia’s Data Catalogue (Government of British Columbia, 2021b). Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested ‘reserved’ areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b). Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.  
Analysis

@@

## Spatial accuracy
        Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks. In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a). Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data. There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b). There are also instances where areas were likely misidentified as cutblocks (Figure 4c). This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.
Figure 5 shows the percentage of matching pixels by category between observed and projected rasters. The percentage of roaded prior to 1990 was the same across all projection methods for the respective metrics because the same 1990 road network was used for all projection as the input for the existing road network. The agree roadless category represented most of the landscape for all projection methods. The subsequent 3 categories (Agree roaded after 1990, project roaded & observe roadless, project roadless & observe roaded) represent the areas where change occurred following 1990.
With the road disturbance footprint metric, the agree roaded after 1990 represented the largest percentage of the landscape which underwent change after 1990. The project roaded & observe roadless category represented higher percentages of the landscape than the project roadless & observe roaded, especially when using a higher sampling density, which shows that there were more areas where the model over projected roads than under projected roads. The small percentage of project roadless & observe roaded across all projection methods shows that there were very few observed roaded areas that the projections did not address.  
Using the forestry disturbance footprint metric, the post-1990 roaded pixels lined up very well. Nearly 70% of all post-1990 projected roaded pixels matched with the observed network when using this metric.  Project roadless & observe roaded very slightly made up a higher percentage of the landscape when compared to project roaded & observe roadless (by less than half a percent across all projection methods). 
When using the fine resolution road presence raster to compare with the observed road presence raster, very few of the post-1990 pixels align. Nearly all post-1990 roaded pixels did not match between observed and projected. The sampling density used determined whether there were more project roaded & observe roadless pixels (higher sampling density) or project roadless & observe roaded pixels (centroid and lower sampling density) amongst the post-1990 roaded pixels.

<!-- Here's some example analysis code: -->

@@

# Discussion
Through assessing the performance of our model by comparing its spatial and aspatial
accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

@@

# Conclusion
The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort. This model, however, is not suitable for strategic planning. The results from our projections in our study area overall matched up adequately with the observed present day forest road networks, with major discrepancies being attributed to issues in the available road and cutblock data. These discrepancies would not be issues in future landscape simulations, only when assessing accuracy. In areas where the quality of the data was high, projections matched well with the observed network. Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations. 
        This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape. With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale.  With land use activities transforming a large proportion of the planet’s land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats. Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of ~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005). With only 7% of the planet’s roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016). The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.

>>>>

(Introduction round 2: road context)
Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use. These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network. These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes. On the other hand, many spatial stochastic forest management simulations haven’t included forest road development at all (citations). One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes. As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).

The goal of our model is to provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations, and can also be applied as an element of more complex models. Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes. Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network. In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially.

In the first section of the paper (methods), we introduce the study area used to assess the accuracy of the projection, we describe the forest road projection methods, the data required, and the various metrics used in order to analyze the similarities and differences between the projected and observed forest road networks across the landscape. Next, we present and discuss the results of our projections across the landscape and compare the metrics spatially and aspatially.

@@

## Road network projection methods
Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We used three methods for solving the MTAP in the roads package. These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST). In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.

LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959). Dijkstra’s algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section “weighted graph”). Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction. This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph. The minimum spanning tree was solved using Kruskal’s algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop. This method was most similar to Clark et al. (2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling. An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.

A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods. All the default settings such as skidding distance, etc. were left as they were within the QGIS plugin.

@@

Table 1. Definitions of road building costs for building the cost surface. 
```{r}
tbl <- "Cost Type|Road Attribute|Value ($ per km)|Source
Barriers of road development|Major lakes|infinity|Major water bodies of BC
Impedance|Existing roads|0.01|Assumed
Road Construction Cost Structure|Pipe crossing|1911|Interior Appraisal Manual
|River crossing|2130|Interior Appraisal Manual
|Base cost|16178 + 504*slope|Interior Appraisal Manual
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df)
```

The road network used was a combination of the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) datasets from the Government of British Columbia’s Data Catalogue. The existing road network inputted into the model was created by filtering out the forest tenure roads with an award date greater than 1990. Because the objective of the roads package is to project forest resource roads, the DRA roads were left untouched for the existing 1990 road network. The projection results would be compared with the observed modern road network which was simply a complete and unfiltered combination of the DRA and Forest Tenure road datasets, representing the most complete and up-to-date road network for the area.   
Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) Government of British Columbia’s Data Catalogue (Government of British Columbia, 2021b). Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested ‘reserved’ areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b). Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.

## Analysis

@@

## Spatial accuracy
  Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks. In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a). Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data. There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b). There are also instances where areas were likely misidentified as cutblocks (Figure 4c). This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.

Figure 5 shows the percentage of matching pixels by category between observed and projected rasters. The percentage of roaded prior to 1990 was the same across all projection methods for the respective metrics because the same 1990 road network was used for all projection as the input for the existing road network. The agree roadless category represented most of the landscape for all projection methods. The subsequent 3 categories (Agree roaded after 1990, project roaded & observe roadless, project roadless & observe roaded) represent the areas where change occurred following 1990.

With the road disturbance footprint metric, the agree roaded after 1990 represented the largest percentage of the landscape which underwent change after 1990. The project roaded & observe roadless category represented higher percentages of the landscape than the project roadless & observe roaded, especially when using a higher sampling density, which shows that there were more areas where the model over projected roads than under projected roads. The small percentage of project roadless & observe roaded across all projection methods shows that there were very few observed roaded areas that the projections did not address.

Using the forestry disturbance footprint metric, the post-1990 roaded pixels lined up very well. Nearly 70% of all post-1990 projected roaded pixels matched with the observed network when using this metric.  Project roadless & observe roaded very slightly made up a higher percentage of the landscape when compared to project roaded & observe roadless (by less than half a percent across all projection methods).

When using the fine resolution road presence raster to compare with the observed road presence raster, very few of the post-1990 pixels align. Nearly all post-1990 roaded pixels did not match between observed and projected. The sampling density used determined whether there were more project roaded & observe roadless pixels (higher sampling density) or project roadless & observe roaded pixels (centroid and lower sampling density) amongst the post-1990 roaded pixels.

@@

# Discussion
Through assessing the performance of our model by comparing its spatial and aspatial
accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

@@

# Conclusion
The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort. This model, however, is not suitable for strategic planning. The results from our projections in our study area overall matched up adequately with the observed present day forest road networks, with major discrepancies being attributed to issues in the available road and cutblock data. These discrepancies would not be issues in future landscape simulations, only when assessing accuracy. In areas where the quality of the data was high, projections matched well with the observed network. Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations.

This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape. With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale.  With land use activities transforming a large proportion of the planet’s land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats. Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of ~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005). With only 7% of the planet’s roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016). The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.
